---
title: "Exploratory time series analysis using R"
author: "1. Wrangling time series data"
toc: true
classoption: aspectratio=169
titlecolor: burntorange
output:
  binb::monash:
    latex_engine: xelatex
    colortheme: monashwhite
    fig_width: 7.5
    fig_height: 3
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
source("setup.R")
```

# Introduction

## Instructor

\placefig{0.6}{1.5}{width=3.5cm}{Rob}
\begin{textblock}{6.95}(5.6,1.3)
\begin{alertblock}{Rob J Hyndman}
\href{https://robjhyndman.com}{\faicon{home} robjhyndman.com}\\
\href{https://aus.social/@robjhyndman}{\faicon{mastodon} @robjhyndman@aus.social}\\
\href{https://github.com/robjhyndman}{\faicon{github}  @robjhyndman}\\
\href{mailto:rob.hyndman@monash.edu}{\faicon{envelope}  rob.hyndman@monash.edu}
\end{alertblock}
\end{textblock}

\vspace*{5cm}\pause

### Helpers

* Nuwani Palihawadana
* Swen Kuh

## Key reference

\begin{block}{}\bf
\hangafter=1\hangindent=.3cm
 {Hyndman, R.~J. \& Athanasopoulos, G. (2021) \emph{Forecasting: principles and practice}, 3rd ed.}
\end{block}\pause
\begin{alertblock}{}\large
\centerline{\bf OTexts.org/fpp3/}
\end{alertblock}

\pause

  * Free and online
  * Data sets in associated R package
  * R code for examples

\pause\vspace*{0.3cm}

### Install required packages

```r
install.packages(c("tidyverse", "fpp3", "GGally"))
```

## Approximate outline

\vspace*{-0.1cm}\centering\fontsize{12}{12}\sf
\begin{tabular}{rp{7cm}l}
  \toprule
  \bf Session & \bf Topic                   & \bf Chapter \\
  \midrule
  1       & Wrangling time series data  & 2 \\
  2       & Visualizing trend and seasonal patterns  & 2, 3 \\
  3       & Compute time series features for large collections of time series  & 4 \\
  \bottomrule
\end{tabular}

\vspace*{.1cm}\begin{alertblock}{}{\centerline{\Large\textbf{bit.ly/etsa2022}}}
\end{alertblock}

# Time series data and tsibbles

## Tidyverts packages

\begin{textblock}{4.2}(8,0)\begin{alertblock}{}\Large\textbf{tidyverts.org}\end{alertblock}\end{textblock}

\placefig{1}{1.4}{width=3.5cm}{tsibble.png}
\placefig{5}{1.4}{width=3.5cm}{tsibbledata.png}
\placefig{3}{4.85}{width=3.5cm}{feasts.png}
\placefig{7}{4.85}{width=3.5cm}{fable.png}

## Time series data

  - Four-yearly Olympic winning times
  - Annual Google profits
  - Quarterly Australian beer production
  - Monthly rainfall
  - Weekly retail sales
  - Daily IBM stock prices
  - Hourly electricity demand
  - 5-minute freeway traffic counts
  - Time-stamped stock transaction data

## Class packages

```{r}
library(fpp3)
```

Loads:

* Data sets
* Some `tidyverse` packages
* `tsibble`, `tsibbledata`, `feasts` and `fable`

## `tsibble` objects

\fontsize{10}{11.2}\sf

```{r, echo = TRUE}
global_economy
```

\only<2->{\begin{textblock}{.75}(2.09,3.68)
\begin{alertblock}{}\fontsize{10}{10}\sf Index\phantom{dg}\end{alertblock}
\end{textblock}}
\only<3->{\begin{textblock}{1.6}(3.30,3.68)
\begin{alertblock}{}\fontsize{10}{10}\sf Key\phantom{dg}\end{alertblock}
\end{textblock}}
\only<4>{\begin{textblock}{6.3}(5.72,3.68)
\begin{alertblock}{}\fontsize{10}{10}\sf Measured variables\phantom{dg}\end{alertblock}
\end{textblock}}

## `tsibble` objects

\fontsize{10}{11.3}\sf

```{r, echo = TRUE}
tourism
```

\only<3->{\begin{textblock}{1.1}(2.1,3.65)
\begin{alertblock}{}\fontsize{10}{10}\sf Index\phantom{dg}\end{alertblock}
\end{textblock}}
\only<4->{\begin{textblock}{3.8}(3.55,3.65)
\begin{alertblock}{}\fontsize{10}{10}\sf Keys\phantom{dg}\end{alertblock}
\end{textblock}}
\only<5>{\begin{textblock}{1.5}(7.85,3.65)
\begin{alertblock}{}\fontsize{10}{10}\sf Measure\phantom{dg}\end{alertblock}
\end{textblock}}

\only<2->{\begin{textblock}{3}(9,5)
\begin{block}{}\fontsize{10}{10}\sf Domestic visitor nights in thousands by state/region and purpose.\phantom{dg}\end{block}
\end{textblock}}

## `tsibble` objects

* A `tsibble` allows storage and manipulation of multiple time series in R.

* It contains:

  + An index: time information about the observation
  + Measured variable(s): numbers of interest
  + Key variable(s): optional unique identifiers for each series

* It works with tidyverse functions.

## The `tsibble` index

\begin{block}{}
For observations more frequent than once per year, we need to use a time class function on the index.
\end{block}
\fontsize{12}{13}\sf

```{r tstablemonth, echo=FALSE}
z <- tibble(Month = paste(2019, month.abb[1:5]), Observation = c(50, 23, 34, 30, 25))
# knitr::kable(z, booktabs=TRUE)
```

```{r tstablemonth2}
z
```

## The `tsibble` index

\begin{block}{}
For observations more frequent than once per year, we need to use a time class function on the index.
\end{block}
\fontsize{12}{13}\sf

```{r month-tsibble}
z |>
  mutate(Month = yearmonth(Month)) |>
  as_tsibble(index = Month)
```

## The `tsibble` index

Common time index variables can be created with these functions:

###
```{r tstable2, echo=FALSE}
tribble(
  ~`Frequency`, ~Function,
  "Quarterly", "`yearquarter()`",
  "Monthly", "`yearmonth()`",
  "Weekly", "`yearweek()`",
  "Daily", "`as_date()`, `ymd()`",
  "Sub-daily", "`as_datetime()`"
) |>
  knitr::kable(booktabs = TRUE)
```

# Example: Australian prison population

## Australian prison population

\full{Beechworth_prison}

## Read a csv file and convert to a tsibble {-}
\fontsize{10}{11}\sf

```{r prison}
prison <- readr::read_csv("data/prison_population.csv")
```
```{r prison2a, dependson="prison", echo=FALSE}
prison
```
## Read a csv file and convert to a tsibble {-}
\fontsize{10}{11}\sf

```{r prison3}
prison <- readr::read_csv("data/prison_population.csv") |>
  mutate(Quarter = yearquarter(date))
```

```{r prison3a, dependson="prison3", echo=FALSE}
prison
```

## Read a csv file and convert to a tsibble {-}
\fontsize{10}{11}\sf

```{r prison4}
prison <- readr::read_csv("data/prison_population.csv") |>
  mutate(Quarter = yearquarter(date)) |>
  select(-date)
```

```{r prison4a, dependson="prison4", echo=FALSE}
prison
```

## Read a csv file and convert to a tsibble {-}
\fontsize{10}{11}\sf

```{r prison5}
prison <- readr::read_csv("data/prison_population.csv") |>
  mutate(Quarter = yearquarter(date)) |>
  select(-date) |>
  as_tsibble(
    index = Quarter,
    key = c(state, gender, legal, indigenous)
  )
```

```{r prison5a, dependson="prison5", echo=FALSE}
prison
```

# Example: Australian pharmaceutical sales

## Australian Pharmaceutical Benefits Scheme

\full{pills}

## Australian Pharmaceutical Benefits Scheme
\begin{block}{}
The \alert{Pharmaceutical Benefits Scheme} (PBS) is the Australian government drugs subsidy scheme.
\end{block}
\pause\fontsize{13.3}{15}\sf

  * Many drugs bought from pharmacies are subsidised to allow more equitable access to modern drugs.
  * The cost to government is determined by the number and types of drugs purchased. Currently nearly 1\% of GDP.
  * The total cost is budgeted based on forecasts of drug usage.
  * Costs are disaggregated by drug type (ATC1 x`r length(unique(PBS$ATC1))` / ATC2 `r length(unique(PBS$ATC2))`), concession category (x`r length(unique(PBS$Concession))`) and patient type (x`r length(unique(PBS$Type))`), giving $84\times2\times2=`r 84*2*2`$ time series.

## Working with `tsibble` objects {-}
\fontsize{8}{10}\sf

```{r pbs1}
PBS
```

## Working with `tsibble` objects {-}
\fontsize{12}{14}\sf

We can use the `filter()` function to select rows.

\fontsize{8}{10}\sf

```{r pbs2}
PBS |>
  filter(ATC2 == "A10")
```

## Working with `tsibble` objects {-}
\fontsize{12}{14}\sf

We can use the `select()` function to select columns.

\fontsize{8}{10}\sf

```{r pbs3}
PBS |>
  filter(ATC2 == "A10") |>
  select(Month, Concession, Type, Cost)
```

## Working with `tsibble` objects {-}
\fontsize{12}{14}\sf

We can use the `summarise()` function to summarise over keys.

\fontsize{8}{10}\sf

```{r pbs4}
PBS |>
  filter(ATC2 == "A10") |>
  select(Month, Concession, Type, Cost) |>
  summarise(total_cost = sum(Cost))
```

## Working with `tsibble` objects {-}
\fontsize{12}{14}\sf

We can use the `mutate()` function to create new variables.

\fontsize{8}{10}\sf

```{r pbs5}
PBS |>
  filter(ATC2 == "A10") |>
  select(Month, Concession, Type, Cost) |>
  summarise(total_cost = sum(Cost)) |>
  mutate(total_cost = total_cost / 1e6)
```

## Working with `tsibble` objects {-}
\fontsize{12}{14}\sf

We can use the `mutate()` function to create new variables.

\fontsize{8}{10}\sf

```{r pbs6}
PBS |>
  filter(ATC2 == "A10") |>
  select(Month, Concession, Type, Cost) |>
  summarise(total_cost = sum(Cost)) |>
  mutate(total_cost = total_cost / 1e6) -> a10
```

```{r a10, echo=FALSE, dependson="pbs6"}
a10
```

# Lab Session 1

## Lab Session 1
\fontsize{14}{15}\sf

 1. Download [`tourism.xlsx`](http://robjhyndman.com/data/tourism.xlsx) from [`http://robjhyndman.com/data/tourism.xlsx`](http://robjhyndman.com/data/tourism.xlsx), and read it into R using `read_excel()` from the `readxl` package.
 2. Create a tsibble which is identical to the `tourism` tsibble from the `tsibble` package.
 3. Find what combination of `Region` and `Purpose` had the maximum number of overnight trips on average.
 4. Create a new tsibble which combines the Purposes and Regions, and just has total trips by State.
