---
title: "Exploratory time series analysis using R"
author: "3. Time series features"
toc: true
classoption: aspectratio=169
titlecolor: burntorange
output:
  binb::monash:
    latex_engine: xelatex
    colortheme: monashwhite
    fig_width: 7.5
    fig_height: 3
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
source("setup.R")
```

# STL Features

## Strength of seasonality and trend

\begin{alertblock}{STL decomposition}
\centerline{$y_t = T_t+S_t+R_t$}
\end{alertblock}

### Seasonal strength
$$\max\left(0, 1-\frac{\text{Var}(R_t)}{\text{Var}(S_t+R_t)}\right)$$

### Trend strength
$$\max\left(0, 1-\frac{\text{Var}(R_t)}{\text{Var}(T_t+R_t)}\right)$$

## Feature extraction and statistics

\fontsize{9}{10}\sf

```{r features}
PBS |> 
  features(Scripts, feat_stl) 
```

## Feature extraction and statistics
\fontsize{8}{9}\sf

```{r features-plot, fig.height=3.8, fig.width=10}
PBS |> features(Scripts, feat_stl) |> 
  ggplot(aes(x = trend_strength, y = seasonal_strength_year, col=ATC1)) +
  geom_point() + facet_grid(Concession ~ Type)
```

## Feature extraction and statistics
\fontsize{9}{9}\sf

Find the most seasonal time series:

```{r extreme}
most_seasonal <- PBS |>
  features(Scripts, feat_stl) |>
  arrange(desc(seasonal_strength_year)) |> 
  head(1)
```

\pause\vspace*{-0.3cm}

```{r extreme2, fig.height=1.6}
PBS |>
  right_join(most_seasonal, by = c("ATC1", "ATC2", "Concession", "Type")) |>
  ggplot(aes(x = Month, y = Scripts)) +
  geom_line() + facet_grid(vars(ATC2, Concession, Type))
```

## Feature extraction and statistics
\fontsize{9}{9}\sf

Find the most trended time series:

```{r extreme3}
most_trended <- PBS |>
  features(Scripts, feat_stl) |>
  arrange(desc(trend_strength)) |> 
  head(1)
```

\pause\vspace*{-0.3cm}

```{r extreme4, fig.height=1.6}
PBS |>
  right_join(most_trended, by = c("ATC1", "ATC2", "Concession", "Type")) |>
  ggplot(aes(x = Month, y = Scripts)) +
  geom_line() + facet_grid(vars(ATC2, Concession, Type))
```

# Dimension reduction for features

```{r pca, echo=FALSE, include=FALSE}
# Save pdf figures
savepdf <- function(file, width = 16, height = 10) {
  fname <<- paste("figs/", file, ".pdf", sep = "")
  pdf(fname, width = width / 2.54, height = height / 2.54, pointsize = 10)
  par(mgp = c(2.2, 0.45, 0), tcl = -0.4, mar = c(3.3, 3.6, 1.1, 1.1))
}
endpdf <- function() {
  crop::dev.off.crop(fname)
}
# Compute features
PBS_features <- PBS |>
  features(Scripts, feature_set(pkgs = "feasts")) |> 
  select(-`...26`) |> 
  na.omit()
# Compute PCs
pcs <- PBS_features |>
  select(-ATC1, -ATC2, -Type, -Concession) |>
  prcomp(scale = TRUE) |>
  broom::augment(PBS_features)
# Save some PC plots
savepdf("pca1", 18, 10)
pcs |> ggplot(aes(x = .fittedPC1, y = .fittedPC2)) +
  geom_point() +
  theme(aspect.ratio = 1)
endpdf()
savepdf("pca2", 18, 10)
pcs |> ggplot(aes(x = .fittedPC1, y = .fittedPC2, col = Concession)) +
  geom_point() +
  theme(aspect.ratio = 1)
endpdf()
savepdf("pca3", 18, 10)
pcs |> ggplot(aes(x = .fittedPC1, y = .fittedPC2, col = Type)) +
  geom_point() +
  theme(aspect.ratio = 1)
endpdf()
savepdf("pca4", 18, 10)
pcs |> ggplot(aes(x = .fittedPC1, y = .fittedPC2, col = ATC1)) +
  geom_point() +
  theme(aspect.ratio = 1)
endpdf()
# Find outliers
outliers <- pcs |>
  filter(.fittedPC2 < -15)
savepdf("pca5", 18, 10)
pcs |> ggplot(aes(x = .fittedPC1, y = .fittedPC2, col=Type)) +
  geom_point() +
  theme(aspect.ratio = 1) +
  geom_point(data = outliers, aes(x = .fittedPC1, y = .fittedPC2), col = "black", shape = 1, size = 3)
endpdf()
```

## Feature extraction and statistics
\fontsize{9}{9}\sf

```{r PBSfeatures, eval=FALSE}
PBS_features <- PBS |>
  features(Scripts, feature_set(pkgs = "feasts")) |> 
  select(-`...26`) |> 
  na.omit()
```

```{r echo=FALSE, cache=FALSE}
PBS_features
```

\begin{textblock}{4}(10.6,1.6)
\begin{alertblock}{}\fontsize{10}{12}\sf
All features from the feasts package
\end{alertblock}
\end{textblock}

## Feature extraction and statistics
\fontsize{9}{8.5}\sf

```{r featurelist, cache=FALSE}
colnames(PBS_features)
```


## Feature extraction and statistics
\fontsize{9}{9}\sf

```{r pcatable}
pcs <- PBS_features |>
  select(-ATC1, -ATC2, -Type, -Concession) |>
  prcomp(scale = TRUE) |>
  broom::augment(PBS_features)
```

```{r echo=FALSE}
pcs
```

\begin{textblock}{4.5}(9.6,1.5)
\begin{alertblock}{}\fontsize{10}{12}\sf
Principal components based on all features from the feasts package
\end{alertblock}
\end{textblock}

## Feature extraction and statistics
\fontsize{9}{9}\sf

\begin{textblock}{3.3}(.4,3)
\begin{alertblock}{}\fontsize{10}{12}\sf
Principal components based on all features from the feasts package
\end{alertblock}
\end{textblock}

```r
pcs |> ggplot(aes(x=.fittedPC1, y=.fittedPC2)) +
  geom_point() + theme(aspect.ratio=1)
```

\placefig{4}{2.6}{height=6.4cm, width=12cm}{pca1}
\vspace*{10cm}

## Feature extraction and statistics
\fontsize{9}{9}\sf

\begin{textblock}{3.3}(.4,3)
\begin{alertblock}{}\fontsize{10}{12}\sf
Principal components based on all features from the feasts package
\end{alertblock}
\end{textblock}

```r
pcs |> ggplot(aes(x=.fittedPC1, y=.fittedPC2, col = ATC1)) +
  geom_point() + theme(aspect.ratio=1)
```

\placefig{4}{2.6}{height=6.4cm, width=12cm}{pca4}
\vspace*{10cm}

## Feature extraction and statistics
\fontsize{9}{9}\sf

\begin{textblock}{3.3}(.4,3)
\begin{alertblock}{}\fontsize{10}{12}\sf
Principal components based on all features from the feasts package
\end{alertblock}
\end{textblock}

```r
pcs |> ggplot(aes(x=.fittedPC1, y=.fittedPC2, col=Concession)) +
  geom_point() + theme(aspect.ratio=1)
```

\placefig{4}{2.6}{height=6.4cm, width=12cm}{pca2}
\vspace*{10cm}

## Feature extraction and statistics
\fontsize{9}{9}\sf

\begin{textblock}{3.3}(.4,3)
\begin{alertblock}{}\fontsize{10}{12}\sf
Principal components based on all features from the feasts package
\end{alertblock}
\end{textblock}

```r
pcs |> ggplot(aes(x=.fittedPC1, y=.fittedPC2, col=Type)) +
  geom_point() + theme(aspect.ratio=1)
```

\only<1>{\placefig{4}{2.6}{height=6.4cm, width=12cm}{pca3}}
\only<2>{\placefig{4}{2.6}{height=6.4cm, width=12cm}{pca5}}
\vspace*{10cm}

## Feature extraction and statistics
\fontsize{8}{8}\sf

```{r outliers2, fig.height=4.5, fig.width=12, out.height="55%"}
filter(pcs, .fittedPC2 < -15) |> 
  left_join(PBS, by = c("ATC1", "ATC2", "Concession", "Type")) |>
  mutate(Series = glue("{ATC2}", "{Concession}", "{Type}", .sep = "\n\n")) |>
  ggplot(aes(x = Month, y = Scripts)) +
  geom_line() + facet_grid(Series ~ .) +
  labs(title = "Outlying time series in PC space")
```

# Lab Session 4

## Lab Session 4

* Find the most seasonal time series in the tourism data.

* Which state has the strongest trends?

* Use a feature-based approach to look for outlying series in `tourism`.

* What is unusual about the series you identify as outliers?
