---
title: "Exploratory time series analysis using R"
author: "2. Visualising trend and seasonality"
toc: true
classoption: aspectratio=169
titlecolor: burntorange
output:
  binb::monash:
    latex_engine: xelatex
    colortheme: monashwhite
    fig_width: 7.5
    fig_height: 3
    includes:
      in_header: header.tex
---

```{r setup, include=FALSE}
source("setup.R")
library(gganimate)
```

# Time plots

## Time plots: `autoplot()`

```{r maxtemp, fig.height=2.4}
maxtemp <- vic_elec |>
  index_by(Day = date(Time)) |>
  summarise(Temperature = max(Temperature))
maxtemp |>
  autoplot(Temperature) +
  labs(x = "Week", y = "Max temperature")
```

## Ansett airlines

\full{ansettlogo}

## Ansett airlines

```{r, echo=TRUE, fig.height=3.1}
ansett |>
  autoplot(Passengers)
```

## Ansett airlines

```{r, echo=TRUE, fig.height=3.1}
ansett |>
  filter(Class == "Economy") |>
  autoplot(Passengers)
```

## Ansett airlines

```{r, echo=TRUE, fig.height=3.1}
ansett |>
  filter(Airports == "MEL-SYD") |>
  autoplot(Passengers)
```

## Ansett airlines

```{r, echo=TRUE, fig.height=3.1}
ansett |>
  filter(Airports == "MEL-SYD") |>
  autoplot(Passengers)
```

\begin{textblock}{2.4}(12.6,7.3)
\begin{alertblock}{}\fontsize{12}{12}\sf
Not the real data! Or is it?
\end{alertblock}
\end{textblock}

# Seasonal plots

## Seasonal plots: `gg_season()`

  * Data plotted against the individual "seasons" in which the data were observed.
  * Something like a time plot except that the data from each season are overlapped.
  * Enables the underlying seasonal pattern to be seen more clearly, and also allows any substantial departures from the seasonal pattern to be easily identified.

## Quarterly Australian Beer Production

```{r, fig.height=2.7}
beer <- aus_production |>
  select(Quarter, Beer) |>
  filter(year(Quarter) >= 1992)
beer |> autoplot(Beer)
```

## Quarterly Australian Beer Production

```{r}
beer |> gg_season(Beer, labels = "right")
```

## Multiple seasonal periods
\fontsize{10}{11}\sf

```{r}
vic_elec
```

## Multiple seasonal periods
\fontsize{12}{13}\sf

```{r, dev = "png", dpi = 180}
vic_elec |> gg_season(Demand)
```

## Multiple seasonal periods
\fontsize{12}{13}\sf

```{r, dev = "png", dpi = 180}
vic_elec |> gg_season(Demand, period = "week")
```

## Multiple seasonal periods
\fontsize{12}{13}\sf

```{r, dev = "png", dpi = 180}
vic_elec |> gg_season(Demand, period = "day")
```

## Seasonal subseries plots: `gg_subseries()`

  * Data for each season collected together in time plot as separate time series.
  * Enables the underlying seasonal pattern to be seen clearly, and changes in seasonality over time to be visualized.

## Quarterly Australian Beer Production

```{r}
beer |> gg_subseries(Beer)
```

# Lab Session 2
## Lab Session 2

Look at the quarterly tourism data for the Snowy Mountains

```r
snowy <- tourism |>
  filter(Region == "Snowy Mountains")
```

- Use `autoplot()`, `gg_season()` and `gg_subseries()` to explore the data.
- What do you learn?

# Time series decompositions

## Time series decomposition

Trend-Cycle
: aperiodic changes in level over time.

Seasonal
: (almost) periodic changes in level due to seasonal factors (e.g., the quarter of the year, the month, or day of the week).

\begin{block}{Additive decomposition}\vspace*{-0.3cm}
\[ y_t = S_t + T_t + R_t \]
\end{block}
\begin{tabular}{@{}llp{8cm}@{}}
where & $y_t=$ & data at period $t$ \\
      & $T_t=$ & trend-cycle component at period $t$\\
      & $S_t=$ & seasonal component at period $t$ \\
      & $R_t=$ & remainder component at period $t$
\end{tabular}


## US Retail Employment
\fontsize{11}{11}\sf

```{r usretail}
us_retail_employment <- us_employment |>
  filter(year(Month) >= 1990, Title == "Retail Trade") |>
  select(-Series_ID)
us_retail_employment
```

\vspace*{10cm}

## US Retail Employment

```{r dable1}
us_retail_employment |>
  autoplot(Employed) +
  labs(y = "Persons (thousands)", title = "Total employment in US retail")
```

\vspace*{10cm}

## US Retail Employment
\fontsize{11}{11}\sf

```{r dable2}
dcmp <- us_retail_employment |>
  model(stl = STL(Employed))
dcmp
```

\pause

*  STL: "Seasonal and Trend decomposition using Loess"

\vspace*{10cm}

## US Retail Employment
\fontsize{11}{11}\sf

```{r dable3}
components(dcmp)
```

\vspace*{10cm}

## US Retail Employment

```{r usretail-stl, fig.width=8, fig.height=5}
components(dcmp) |> autoplot()
```

## US Retail Employment

```{r dable4, fig.height=2.7}
us_retail_employment |>
  autoplot(Employed, color = "gray") +
  autolayer(components(dcmp), trend, color = "#D55E00") +
  labs(y = "Persons (thousands)", title = "Total employment in US retail")
```

\vspace*{10cm}


## US Retail Employment

```{r usretail-sa, fig.height=2.7}
us_retail_employment |>
  autoplot(Employed, color = "gray") +
  autolayer(components(dcmp), season_adjust, color = "#0072B2") +
  labs(y = "Persons (thousands)", title = "Total employment in US retail")
```

\only<2>{\begin{textblock}{5.8}(9.8,6.8)
\begin{block}{Seasonally adjusted data}
$$y_t - S_t$$
\end{block}
\end{textblock}
}

## STL decomposition

```{r stlwindowanim, echo=FALSE, warning=FALSE, message=FALSE, fig.show='animate', interval=1/10,  fig.height=5.35, fig.width=8, aniopts='controls,buttonsize=0.3cm,width=11.5cm', eval=TRUE}
s_windows <- seq(5, 55, by = 2)
stl_defs <- purrr::map(s_windows, function(s_window) {
  STL(Employed ~ season(window = s_window), robust = TRUE)
})
names(stl_defs) <- sprintf("season(window=%02d)", s_windows)

us_retail_employment |>
  model(!!!stl_defs) |>
  components() |>
  as_tibble() |>
  pivot_longer(Employed:remainder,
    names_to = "component", names_ptypes = list(component = factor(levels = c("Employed", "trend", "season_year", "remainder"))),
    values_to = "Employed"
  ) |>
  ggplot(aes(x = Month, y = Employed)) +
  geom_line() +
  facet_grid(rows = vars(component), scales = "free_y") +
  labs(
    title = "STL decomposition of US retail employment",
    subtitle = "{closest_state}"
  ) +
  transition_states(.model)
```

\vspace*{10cm}

## STL decomposition

```{r echo = TRUE, results = 'hide'}
us_retail_employment |>
  model(STL(Employed ~ trend(window = 21) + season(window = 13),
    robust = TRUE
  )) |>
  components()
```

\fontsize{12}{13}\sf

  *  `trend(window = ?)` controls wiggliness of trend component.
  *  `season(window = ?)` controls variation on seasonal component.
  *  `season(window = 'periodic')` is equivalent to an infinite window.

## STL decomposition
\fontsize{13}{15}\sf

* Algorithm that updates trend and seasonal components iteratively.
* Starts with $\hat{T}_t=0$
* Uses a mixture of loess and moving averages to successively refine the trend and seasonal estimates.
* trend window controls loess bandwidth on deseasonalised values.
* season window controls loess bandwidth on detrended subseries.
* Robustness weights based on remainder.
* Default season: `window = 13`
* Default trend: `window =` \newline\mbox{}\hfill `nextodd(ceiling((1.5*period)/(1-(1.5/s.window)))`
* `window` values should be odd numbers for symmetry.

# Lab Session 3

## Lab Session 3
\fontsize{13}{14}\sf

1. Produce an STL decomposition of the Snowy Mountains data.

2. Experiment with different values of the two `window` arguments.

3. Plot the seasonally adjusted series.
